GPU Texture Sampler Bezier Curve Evaluation

### 2016-04-28 Paper/paper.tex
Taking out jcgt for self publishing

\setciteauthor{Alan Wolfe}
\setcitetitle{GPU Efficient Texture Based Bezier Curve Evaluation}

@@

% Once an article is accepted accepted, switch to the following line and comment the preceding one. The editor will supply the argument values.
\accepted{2014-02-07}{2014-02-07}{2014-02-07}{Editor Name}{3}{1}{1}{1}{2014}
\seturl{http://jcgt.org/published/0003/01/01/}

@@

\title{GPU Efficient Texture Based Bezier Curve Evaluation}

\author
       {Alan Wolfe\\Blizzard Entertainment}

@@

The core of this technique is that the $N$-dimensional linear texture interpolation capabilities on the GPU can be mathematically equivalent to De Casteljeau's algorithm for evaluating Bezier curves.  We can use linear texture sampling on textures of increasing dimensionalities to have the texture sampler calculate points on curves of increasing degree.

@@

One dimensional linear texture sampling is just the linear interpolation between two data points.  This is equivalent to the De Casteljeau algorithm when evaluating a linear curve of degree 1.  In both cases, it is just linear interpolation between two values.

@@

The De Casteljeau algorithm for evaluating a quadratic curve (degree 2) is also just a linear interpolation between two linear interpolations.  If the control points of the quadratic curve are $P_0,P_1,P_2$, let's call the first interpolated point $Q_{0,1}$ which is the interpolation between $P_0$ and $P_1$ at $t$. The second linear interpolation is $Q_{1,2}$ and is the interpolation between $P_1$ and $P_2$ at $t$.  The final point on the curve will be the interpolation at $t$ between $Q_{0,1}$ and $Q_{1,2}$, which we can call $R_{0,1,2}$.

@@

To further see how bilinear interpolation can be equivalent to evaluating a quadratic Bezier curve with the De Casteljeau algorithm, \autoref{fig:decasteljeauBilinear} shows it visually, and you can also compare the GLSL source code in \autoref{lst:GLSLCurves}.

\begin{lstlisting}[caption={GLSL implementation of bilinear interpolation and the De Casteljeau algorithm for a quadratic Bezier curve.}, label={lst:GLSLCurves}]
float QuadraticBezier (
  in float t,
  in float P0, in float P1, in float P2
) {
    return mix(mix(P0,P1,t), mix(P1,P2,t), t);
}

@@

  \begin{figure}
    % bezier curve diagram
    \begin{tikzpicture}[x=4cm,y=4cm]
      % border
      \draw (-0.2,-0.2) -- (1.2,-0.2) -- (1.2,1.2) -- (-0.2,1.2) -- (-0.2,-0.2);
      % 1d quadratic bezier curve with control points 0,1,0
      \draw[scale=1,domain=0:1,smooth,variable=\x,blue,line width=0.5mm] plot ({\x},{2*\x-2*\x*\x});    
      % control polygon / labels
      \draw[red,->]       (0.0,0.0) -- (0.5,1.0);
      \draw[darkgreen,->] (0.5,1.0) -- (1.0,0.0);
      \draw[red]          (0.25,0.5) node[anchor=east] {$Q_{0,1}$};
      \draw[darkgreen]    (0.75,0.5) node[anchor=west] {$Q_{1,2}$};
      % control point labels
      \draw (0.0,0.0) node[anchor=east]  {$\bf{P_0}$};
      \draw (0.5,1.0) node[anchor=south] {$\bf{P_1}$};
      \draw (1.0,0.0) node[anchor=west]  {$\bf{P_2}$};
      % draw a line from AB->AC for time = 0.75
      \draw[orange,->] (0.3,0.6) -- (0.8,0.4);
      \draw[orange] (0.40,0.55) node[anchor=south west] {$R_{0,1,2}$};
      % draw the point at t=0.6 and label underneath the curve
      \draw[black,fill=black] (0.6,0.48) circle (1mm);
      \draw[black] (0.5,0.45) node[anchor=north] {\bf{t}=0.6};
    \end{tikzpicture}
    % bilinear diagram
    \begin{tikzpicture}[x=4cm,y=4cm]
        % border
        \draw (-0.2,-0.2) -- (1.2,-0.2) -- (1.2,1.2) -- (-0.2,1.2) -- (-0.2,-0.2);
        % box
        \draw (0,0) -- (1,0) -- (1,1) -- (0,1) -- (0,0);
        % representative line for where the quadratic bezier curve lives
        \draw[blue,line width=0.5mm] (0,0) -- (1,1);    
        % corner labels
        \draw (0,0) node[anchor=north east] {$\bf{P_0}$};
        \draw (1,0) node[anchor=north west] {$\bf{P_1}$};     
        \draw (0,1) node[anchor=south east] {$\bf{P_1}$};
        \draw (1,1) node[anchor=south west] {$\bf{P_2}$};     
        % x axis interpolation
        \draw[red,->]       (0.0,0.1) -- (1.0,0.1);
        \draw[red]          (0.4,0.1) node[anchor=south] {$Q_{0,1}$};
        \draw[darkgreen,->] (0.0,0.9) -- (1.0,0.9);
        \draw[darkgreen]    (0.4,0.9) node[anchor=north] {$Q_{1,2}$};
        % y axis interpolation
        \draw[orange,->]    (0.6,0.0) -- (0.6,1.0);
        \draw[orange]     (0.6,0.5) node[anchor=west] {$R_{0,1,2}$};
        % draw the point at t=0.6 and label for it
        \draw[black,fill=black] (0.6,0.6) circle (1mm);
        \draw[black] (0.55,0.6) node[anchor=east] {\bf{t}=0.6};
    \end{tikzpicture} 
    \caption{\textit{Left:} A quadratic Bezier curve evaluated at {\bf{t}}=0.6 using the De Casteljeau algorithm.  \textit{Right:} The De Casteljeau algorithm using bilinear interpolation.}
    \label{fig:decasteljeauBilinear}
  \end{figure}

The pattern continues for higher texture dimensions and curve degrees as well.  Linear texture sampling of textures of dimension $N$ is just the linear interpolation between two linear texture samples of dimension $N-1$.  Using the De Casteljeau algorithm to evaluate a curve of degree $N$ just means that you take a linear interpolation between curves of degree $N-1$.

Compared to a quadratic curve evaluated with De Casteljeau's algorithm, bilinear texture sampling has two blend weights instead of one, and has four values to interpolate between instead of three, but you can set a texture up such that when you sample it a specific way, you get as output a quadratic Bezier curve.  Because of this, the De Casteljeau algorithm is a subset of the $N$-dimensional linear texture sampling available on the GPU.

@@

It is also possible to get points on curves of higher degree by taking multiple texture reads and combining them either using the De Casteljeau algorithm, or by using the Bernstein form equation of Bezier curves.

@@

The De Casteljeau algorithm is equivalent to the Bernstein form of Bezier curves (\autoref{eqn:bernsteinform}).

@@

      % Comments for each level
      \begin{scope}[every node/.style={right}]
        \path (Root      -| Root-2-2-2) ++(5mm,0) node {$\Rightarrow$} ++(5mm,0) node [blue] {Cubic / Trilinear};
        \path (Root-1    -| Root-2-2-2) ++(5mm,0) node {$\Rightarrow$} ++(5mm,0) node [blue] {Quadratic / Bilinear};
        \path (Root-1-1  -| Root-2-2-2) ++(5mm,0) node {$\Rightarrow$} ++(5mm,0) node [blue] {Linear / Linear};
        \path (Root-1-1-1-| Root-2-2-2) ++(5mm,0) node {$\Rightarrow$} ++(5mm,0) node [blue] {Point / Nearest};
      \end{scope}
    \end{scope}   
  \end{tikzpicture}
  \caption{A tree showing the De Casteljeau algorithm for a cubic curve.  The labels on the right show what type of curves are evaluated at that level, as well as the $N$ dimensional sampling that is required to evaluate nodes at that level with a single texture read.}
  \label{fig:decdimensionality}
\end{figure}

@@

Since the De Casteljeau algorithm just interpolates between linear interpolations, we can take more texture samples and combine them in the shader program to get higher degree curves.  For a degree $N$ Bezier curve, the De Casteljeau algorithm requires $2^{N-1}$ linear interpolations for the linear level of the tree.  You need two pixels to perform a linear interpolation, so it seems that we would need $2^N$ pixels to store a curve of degree $N$ in a 1 dimensional textures.  For a cubic curve, that would look like this: $[P_0,P_1,P_1,P_2,P_1,P_2,P_2,P_3]$.

@@

Taking $N$ linearly interpolated texture samples from a one dimensional texture of size $N+1$ gives us the linear interpolated values we would need to continue to perform the De Casteljeau algorithm.  We can compute the final curve point either by doing that, or by plugging the values into the Bernstein form equation of Bezier curves.

In the case of a cubic curve with four control points, the one dimensional texture will be 4 pixels in size, and three texture reads will need to be done to get the interpolations along the line segments for $\overline{P_0P_1},\overline{P_1P_2},\overline{P_2P_3}$.  Those will then need to be combined using either the De Casteljeau algorithm, or by using the Bernstein form of a quadratic Bezier curve ($P = P_0*s^2 + P_1*2st + P_2*t^2$) to elevate those three points from degree 1 to degree 3, to get the final point on the curve.  Sample code of this example is provided in \autoref{lst:GLSLCubicTexture1D}.

\begin{lstlisting}[caption={GLSL for evaluating a cubic curve encoded in a 4 pixel 1d texture.  Linear texture sampling is used to evaluate the linear level of the De Casteljeau algorithm, then the process is continued both with the De Casteljeau algorithm, as well as the Bernstein form of a quadratic Bezier curve.}, label={lst:GLSLCubicTexture1D}]
uniform sampler1D uSampler;  
const float c_textureSize = 4.0; // $P_0,P_1,P_2,P_3$

@@

\begin{lstlisting}[caption={GLSL for evaluating a cubic curve encoded in a $(2,4)$ pixel 2d texture.  Bilinear texture sampling used to evaluate the first two levels of the De Casteljeau algorithm, then the process is continued both with the De Casteljeau algorithm, as well as the Bernstein form of a linear Bezier curve (lerp).}, label={lst:GLSLCubicTexture2D}]
uniform sampler2D uSampler; 
const vec2 c_textureSize = vec2(2.0,4.0); // rounded up from 2x3

@@

\begin{lstlisting}[caption={GLSL for evaluating a cubic curve encoded in a $(2,2,2)$ pixel 3d texture.  Trilinear texture sampling used to evaluate all three levels of the De Casteljeau algorithm.}, label={lst:GLSLCubicTexture3D}]
// 2x2x2 3d texture.
uniform sampler2D uSampler; 
const vec3 c_textureSize = vec3(2.0, 2.0, 2.0);

@@

You could also set it up such that R,G,B,A were meant to be combined together to make a quintic (degree 5) curve.  You'd only get a single curve value per texture read, but it would be 3 degrees higher.  You could combine the color channels either with De Casteljeau's algorithm, or with the cubic Bernstein Bezier form.

@@

A better approach that gives the same results would be to only do one nearest neighbor texture sample per control point in the texture, and then in the shader program, either use De Casteljeau's algorithm to combine them, or use the Bernstein form equation of Bezier curves.  At this point, you are best off just using a 1D texture to store your control points in, since you don't need any of the built in interpolation.

@@

\hspace{-2mm}\begin{tabular}{p{0.5\textwidth}p{0.5\textwidth}}
Alan Wolfe \newline
Blizzard Entertainment \newline
16215 Alton Parkway \newline
Irvine, CA 92618 \newline
\href{mailto:awolfe@blizzard.com}{awolfe@blizzard.com}
\href{http://blog.demofox.org}{http://blog.demofox.org}
\end{tabular}

>>>>

\setciteauthor{Alan Wolfe}
\setcitetitle{GPU Texture Sampler Bezier Curve Evaluation}

@@

% Once an article is accepted accepted, switch to the following line and comment the preceding one. The editor will supply the argument values.
\accepted{2014-02-07}{2014-02-07}{2014-02-07}{Editor Name}{3}{1}{1}{1}{2014}
\seturl{http://blog.demofox.org/}

@@

\title{GPU Texture Sampler Bezier Curve Evaluation}

\author
       {Alan Wolfe\\http://blog.demofox.org}

@@

The core of this technique is that the $N$-dimensional linear texture interpolation capabilities on the GPU can be mathematically equivalent to De Casteljau's algorithm for evaluating Bezier curves.  We can use linear texture sampling on textures of increasing dimensionalities to have the texture sampler calculate points on curves of increasing degree.

@@

One dimensional linear texture sampling is just the linear interpolation between two data points.  This is equivalent to the De Casteljau algorithm when evaluating a linear curve of degree 1.  In both cases, it is just linear interpolation between two values.

@@

The De Casteljau algorithm for evaluating a quadratic curve (degree 2) is also just a linear interpolation between two linear interpolations.  If the control points of the quadratic curve are $P_0,P_1,P_2$, let's call the first interpolated point $Q_{0,1}$ which is the interpolation between $P_0$ and $P_1$ at $t$. The second linear interpolation is $Q_{1,2}$ and is the interpolation between $P_1$ and $P_2$ at $t$.  The final point on the curve will be the interpolation at $t$ between $Q_{0,1}$ and $Q_{1,2}$, which we can call $R_{0,1,2}$.

@@

To further see how bilinear interpolation can be equivalent to evaluating a quadratic Bezier curve with the De Casteljau algorithm, \autoref{fig:decasteljeauBilinear} shows it visually, and you can also compare the GLSL source code in \autoref{lst:GLSLCurves}.

\begin{lstlisting}[caption={GLSL implementation of bilinear interpolation and the De Casteljau algorithm for a quadratic Bezier curve.}, label={lst:GLSLCurves}]
float QuadraticBezier (
  in float t,
  in float P0, in float P1, in float P2
) {
    return mix(mix(P0,P1,t), mix(P1,P2,t), t);
}

@@

  \begin{figure}
    % bezier curve diagram
    \begin{tikzpicture}[x=4cm,y=4cm]
      % border
      \draw (-0.2,-0.2) -- (1.2,-0.2) -- (1.2,1.2) -- (-0.2,1.2) -- (-0.2,-0.2);
      % 1d quadratic bezier curve with control points 0,1,0
      \draw[scale=1,domain=0:1,smooth,variable=\x,blue,line width=0.5mm] plot ({\x},{2*\x-2*\x*\x});    
      % control polygon / labels
      \draw[red,->]       (0.0,0.0) -- (0.5,1.0);
      \draw[darkgreen,->] (0.5,1.0) -- (1.0,0.0);
      \draw[red]          (0.25,0.5) node[anchor=east] {$Q_{0,1}$};
      \draw[darkgreen]    (0.75,0.5) node[anchor=west] {$Q_{1,2}$};
      % control point labels
      \draw (0.0,0.0) node[anchor=east]  {$\bf{P_0}$};
      \draw (0.5,1.0) node[anchor=south] {$\bf{P_1}$};
      \draw (1.0,0.0) node[anchor=west]  {$\bf{P_2}$};
      % draw a line from AB->AC for time = 0.75
      \draw[orange,->] (0.3,0.6) -- (0.8,0.4);
      \draw[orange] (0.40,0.55) node[anchor=south west] {$R_{0,1,2}$};
      % draw the point at t=0.6 and label underneath the curve
      \draw[black,fill=black] (0.6,0.48) circle (1mm);
      \draw[black] (0.5,0.45) node[anchor=north] {\bf{t}=0.6};
    \end{tikzpicture}
    % bilinear diagram
    \begin{tikzpicture}[x=4cm,y=4cm]
        % border
        \draw (-0.2,-0.2) -- (1.2,-0.2) -- (1.2,1.2) -- (-0.2,1.2) -- (-0.2,-0.2);
        % box
        \draw (0,0) -- (1,0) -- (1,1) -- (0,1) -- (0,0);
        % representative line for where the quadratic bezier curve lives
        \draw[blue,line width=0.5mm] (0,0) -- (1,1);    
        % corner labels
        \draw (0,0) node[anchor=north east] {$\bf{P_0}$};
        \draw (1,0) node[anchor=north west] {$\bf{P_1}$};     
        \draw (0,1) node[anchor=south east] {$\bf{P_1}$};
        \draw (1,1) node[anchor=south west] {$\bf{P_2}$};     
        % x axis interpolation
        \draw[red,->]       (0.0,0.1) -- (1.0,0.1);
        \draw[red]          (0.4,0.1) node[anchor=south] {$Q_{0,1}$};
        \draw[darkgreen,->] (0.0,0.9) -- (1.0,0.9);
        \draw[darkgreen]    (0.4,0.9) node[anchor=north] {$Q_{1,2}$};
        % y axis interpolation
        \draw[orange,->]    (0.6,0.0) -- (0.6,1.0);
        \draw[orange]     (0.6,0.5) node[anchor=west] {$R_{0,1,2}$};
        % draw the point at t=0.6 and label for it
        \draw[black,fill=black] (0.6,0.6) circle (1mm);
        \draw[black] (0.55,0.6) node[anchor=east] {\bf{t}=0.6};
    \end{tikzpicture} 
    \caption{\textit{Left:} A quadratic Bezier curve evaluated at {\bf{t}}=0.6 using the De Casteljau algorithm.  \textit{Right:} The De Casteljau algorithm using bilinear interpolation.}
    \label{fig:decasteljeauBilinear}
  \end{figure}

The pattern continues for higher texture dimensions and curve degrees as well.  Linear texture sampling of textures of dimension $N$ is just the linear interpolation between two linear texture samples of dimension $N-1$.  Using the De Casteljau algorithm to evaluate a curve of degree $N$ just means that you take a linear interpolation between curves of degree $N-1$.

Compared to a quadratic curve evaluated with De Casteljau's algorithm, bilinear texture sampling has two blend weights instead of one, and has four values to interpolate between instead of three, but you can set a texture up such that when you sample it a specific way, you get as output a quadratic Bezier curve.  Because of this, the De Casteljau algorithm is a subset of the $N$-dimensional linear texture sampling available on the GPU.

@@

It is also possible to get points on curves of higher degree by taking multiple texture reads and combining them either using the De Casteljau algorithm, or by using the Bernstein form equation of Bezier curves.

@@

The De Casteljau algorithm is equivalent to the Bernstein form of Bezier curves (\autoref{eqn:bernsteinform}).

@@

      % Comments for each level
      \begin{scope}[every node/.style={right}]
        \path (Root      -| Root-2-2-2) ++(5mm,0) node {$\Rightarrow$} ++(5mm,0) node [blue] {Cubic / Trilinear};
        \path (Root-1    -| Root-2-2-2) ++(5mm,0) node {$\Rightarrow$} ++(5mm,0) node [blue] {Quadratic / Bilinear};
        \path (Root-1-1  -| Root-2-2-2) ++(5mm,0) node {$\Rightarrow$} ++(5mm,0) node [blue] {Linear / Linear};
        \path (Root-1-1-1-| Root-2-2-2) ++(5mm,0) node {$\Rightarrow$} ++(5mm,0) node [blue] {Point / Nearest};
      \end{scope}
    \end{scope}   
  \end{tikzpicture}
  \caption{A tree showing the De Casteljau algorithm for a cubic curve.  The labels on the right show what type of curves are evaluated at that level, as well as the $N$ dimensional sampling that is required to evaluate nodes at that level with a single texture read.}
  \label{fig:decdimensionality}
\end{figure}

@@

Since the De Casteljau algorithm just interpolates between linear interpolations, we can take more texture samples and combine them in the shader program to get higher degree curves.  For a degree $N$ Bezier curve, the De Casteljau algorithm requires $2^{N-1}$ linear interpolations for the linear level of the tree.  You need two pixels to perform a linear interpolation, so it seems that we would need $2^N$ pixels to store a curve of degree $N$ in a 1 dimensional textures.  For a cubic curve, that would look like this: $[P_0,P_1,P_1,P_2,P_1,P_2,P_2,P_3]$.

@@

Taking $N$ linearly interpolated texture samples from a one dimensional texture of size $N+1$ gives us the linear interpolated values we would need to continue to perform the De Casteljau algorithm.  We can compute the final curve point either by doing that, or by plugging the values into the Bernstein form equation of Bezier curves.

In the case of a cubic curve with four control points, the one dimensional texture will be 4 pixels in size, and three texture reads will need to be done to get the interpolations along the line segments for $\overline{P_0P_1},\overline{P_1P_2},\overline{P_2P_3}$.  Those will then need to be combined using either the De Casteljau algorithm, or by using the Bernstein form of a quadratic Bezier curve ($P = P_0*s^2 + P_1*2st + P_2*t^2$) to elevate those three points from degree 1 to degree 3, to get the final point on the curve.  Sample code of this example is provided in \autoref{lst:GLSLCubicTexture1D}.

\begin{lstlisting}[caption={GLSL for evaluating a cubic curve encoded in a 4 pixel 1d texture.  Linear texture sampling is used to evaluate the linear level of the De Casteljau algorithm, then the process is continued both with the De Casteljau algorithm, as well as the Bernstein form of a quadratic Bezier curve.}, label={lst:GLSLCubicTexture1D}]
uniform sampler1D uSampler;  
const float c_textureSize = 4.0; // $P_0,P_1,P_2,P_3$

@@

\begin{lstlisting}[caption={GLSL for evaluating a cubic curve encoded in a $(2,4)$ pixel 2d texture.  Bilinear texture sampling used to evaluate the first two levels of the De Casteljau algorithm, then the process is continued both with the De Casteljau algorithm, as well as the Bernstein form of a linear Bezier curve (lerp).}, label={lst:GLSLCubicTexture2D}]
uniform sampler2D uSampler; 
const vec2 c_textureSize = vec2(2.0,4.0); // rounded up from 2x3

@@

\begin{lstlisting}[caption={GLSL for evaluating a cubic curve encoded in a $(2,2,2)$ pixel 3d texture.  Trilinear texture sampling used to evaluate all three levels of the De Casteljau algorithm.}, label={lst:GLSLCubicTexture3D}]
// 2x2x2 3d texture.
uniform sampler2D uSampler; 
const vec3 c_textureSize = vec3(2.0, 2.0, 2.0);

@@

You could also set it up such that R,G,B,A were meant to be combined together to make a quintic (degree 5) curve.  You'd only get a single curve value per texture read, but it would be 3 degrees higher.  You could combine the color channels either with De Casteljau's algorithm, or with the cubic Bernstein Bezier form.

@@

A better approach that gives the same results would be to only do one nearest neighbor texture sample per control point in the texture, and then in the shader program, either use De Casteljau's algorithm to combine them, or use the Bernstein form equation of Bezier curves.  At this point, you are best off just using a 1D texture to store your control points in, since you don't need any of the built in interpolation.

@@

\hspace{-2mm}\begin{tabular}{p{0.5\textwidth}p{0.5\textwidth}}
Alan Wolfe \newline
\href{mailto:alan.wolfe@gmail.com}{alan.wolfe@gmail.com}
\href{http://blog.demofox.org}{http://blog.demofox.org}
\end{tabular}
